# `AnnotationHub` for proteomics

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r env, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
suppressPackageStartupMessages(library("BiocStyle"))
```

The aim of this package is to offer access to mass spectrometry and
proteomics data throught the `r Biocpkg("AnnotationHub")`
infrastructure.



```{r ahinit}
library("AnnotationHub")
ah <- AnnotationHub()
ah
```

Let's start by querying the entries that originate from the PRIDE
database:

```{r provider}
query(ah, "PRIDE")
```

Or those of a specific project

```{r title}
ah[grep("PXD000001", ah$title)]
```

To see the metadata of a specific entry, we use its AnnotationHub
entry number inside single `[`

```{r ah49008}
ah["AH49008"]
```

To access the actual data, raw mass spectrometry data in this case, we
double the `[[`

```{r rawmsdata}
library("mzR")
rw <- ah[["AH49008"]]
rw
```

In this case, we have an instance of class `r as.character(class(rw))`,
that can be processed as anticipated.

In the short demonstration above, we had **direct** and
**standardised** access to the raw data, without a need to manually
open this raw data or worry about the file format. The data was
prepared and converted into a **standard Bioconductor data types** for
immediate consumption by the user. This is also valid for other
relevant data types such as identification results, fasta files or
protein of peptide quantitation data.

See the `ProteomicsAnnotationHubData` vignette for details on
available data and how to add new proteomics data to AnnotationHub.
