---
title: "Proteomics Data in Annotation Hub"
author: "Laurent Gatto"
date: "`r doc_date()`"
output:
  BiocStyle::html_document:
   toc_depth: 1
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Proteomics Data in Annotation Hub}
%\VignetteKeyword{proteomics, mass spectrometry, data}
%\VignettePackage{ProteomicsAnnotationHubData}
-->

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

# Introduction



### Updates {-}

To suggest updates and/or new mass spectrometry and/or proteomics
data, please post your suggestions/request on the
[Bioconductor support site](https://support.bioconductor.org/) or open
a
[github issue](https://github.com/lgatto/ProteomicsAnnotationHubData/issues).
Contributions can also be made to the
[package github repository](https://github.com/lgatto/ProteomicsAnnotationHubData).

# Annotation

### Tags {-}

A list of suggested tags is shown below. These suggestions will be
updated and completed over time.

```{r tags, echo=FALSE}
ProteomicsAnnotationHubData:::ProteomicsAnnotationHubDataTags
```

### Providers  {-}

A list of predefined/tested providers.

```{r providers, echo=FALSE}
ProteomicsAnnotationHubData:::ProteomicsAnnotationHubDataProviders
```

### Location prefix {-}

- `.prideBaseUrl` for files on the PRIDE ftp server.
- `.amazonBaseUrl` for files on `AnnotationHub` S3 instance.

### Source types {-}

|---------------|---------|---------|-----------|--------------|--------|
| SourceType    |  mzML   |   mzTab |   mzid    |      FASTA   | MSnSet |
| DispatchClass | mzRpwiz |  MSnSet |  mzRident |  AAStringSet | MSnSet |
| RDataClass    | mzRpwiz |  MSnSet |  mzRident |  AAStringSet | MSnSet |

### Preparer functions {-}

Preparer functions and recipes are only required if the `rda` file is
prepared on the `AnnotationHub` Amazon S3 instance.


# Available datasets

## PXD000001

Description

> Expected reporter ion ratios: Erwinia peptides: 1:1:1:1:1:1 Enolase
> spike (sp|P00924|ENO1_YEAST): 10:5:2.5:1:2.5:10 BSA spike
> (sp|P02769|ALBU_BOVIN): 1:2.5:5:10:5:1 PhosB spike
> (sp|P00489|PYGM_RABIT): 2:2:2:2:1:1 Cytochrome C spike
> (sp|P62894|CYC_BOVIN): 1:1:1:1:1:2.

Four data files from the PRIDE
[PXD000001](http://www.ebi.ac.uk/pride/archive/projects/PXD000001)
experiment are served through `AnnotationHub`.

1. The raw mass spectrometry data, via the
   `TMT_Erwinia_1uLSike_Top10HCD_isol2_45stepped_60min_01-20141210.mzML`
   file from the PRDIE ftp site, served as an `mzRpwiz` object, from
   the `r Biocpkg("mzR")` package.

2. The `mzTab` file, via the `F063721.dat-mztab.txt` file from the
   PRIDE ftp site, served as an `MSnSet` object, from the
   `r Biocpkg("MSnbase")` package.

3. The protein data base, via the `erwinia_carotovora.fasta` file from
	the PRIDE ftp server, served as a `AAStringSet` object, from the
	`r Biocpkg("Biostrings")` package.

4. The identification results, produced using the `MSGF+` search
   engine, served as a `mzRident` object, from the `r Biocpkg("mzR")`
   package.
